### 2020-07-23 code/figure2.Rmd
updated figure2 to include figure 3. removed figure 3 code from manuscript

<div style="text-align: right"> Figure 2 Data </div>
```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

@@

```{r figure-2-plot, fig.width=72, fig.height=48, fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions. Waterfall plot; columns represent tumors. Each row represents a pathway. The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively. All 377 pathways are shown. Red signifies pathway disrupted pathways, grey signifies not disrupted."}
## top100 pathways pancancer oncoprint
load("../data/onco_ggplot.Rda") ## load data
onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint

fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
  geom_tile(aes(fill = as.character(mutation_status))) +
  scale_fill_manual(values = onco_color) +
  xlab(label = "sample_id") +
  labs(tag = "A") +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none")

load("../barplot_pathways_fig2b.rda")
fig2B <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
  geom_col() +
  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
  ylab("Proportion of samples mutated in pathway") +
  xlab("Pathways enriched with mutations within clusters") +
  theme_classic() +
  scale_fill_manual(values = bp.cluster.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.ticks = element_blank(),
    strip.text.y = element_text(size = 12, hjust = .5),
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_blank(),
    legend.position = "none"
  )

fig2B_rotate <- pathway_counts %>% ggplot(aes(y = pathway, x = proportion, fill = color)) +
  geom_col() +
  facet_wrap(~clust_knn, nrow = 1, strip.position = "top") +
  xlab("Proportion of samples mutated in pathway") +
  ylab("Pathways enriched with mutations within clusters") +
  theme_classic() +
  scale_fill_manual(values = bp.cluster.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.ticks = element_blank(),
    strip.text.x = element_text(size = 12, hjust = .5),
    strip.background = element_blank(),
    legend.position = "none"
  )

## top10 pathways for each cluster
load("../data/onco_clust.Rda") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
for(n in 1:length(clust_col_order)) {
  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
  rownames_to_column(var = "path_name") %>% ## path_name to col
  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
  geom_tile(aes(fill = as.character(mutation_status))) +
  scale_fill_manual(values = onco_color) +
  xlab(label = paste0("Class ", n)) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none") 
}

## setting patchwork panel layout
layout <- '
AABCD
AABCD
AABCD
AAEFG
AAEFG
AAEFG
AAHIJ
AAHIJ
AAHIJ
AAKLM
AAKLM
AAKLM
'

fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
```

>>>>

<div style="text-align: right"> Figure 2 Data </div>
```{r}
library(ggplot2)
library(patchwork)
library(tidyverse)
library(cowplot)
```

```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

@@

```{r figure-2-plot, out.height='90%', out.width='90%',fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions. Waterfall plot; columns represent tumors. Each row represents a pathway. The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively. All 377 pathways are shown. Red signifies pathway disrupted pathways, grey signifies not disrupted."}

## top100 pathways pancancer oncoprint
load("../data/onco_ggplot.Rda") ## load data
onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint

fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
  geom_tile(aes(fill = as.character(mutation_status))) +
  scale_fill_manual(values = onco_color) +
  theme_classic() +
  xlab("Pan-Cancer (n=7,607 tumor samples)") +
  ylab("377 pathways") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        #plot.tag = element_text(size = 8, hjust = 0, vjust = 0),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none")

load("../data/effect_heatmap.Rda") ## load data
effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme

guide_axis_label_trans <- function(label_trans = identity, ...) {
  axis_guide <- guide_axis(...)
  axis_guide$label_trans <- rlang::as_function(label_trans)
  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
  axis_guide
}

guide_train.guide_axis_trans <- function(x, ...) {
  trained <- NextMethod()
  trained$key$.label <- x$label_trans(trained$key$.label)
  trained
}

fig2B <- ggplot(data = effect_heatmap, aes(y = cluster, x = path_name)) +
  geom_tile(size = 1, aes(fill = effect_size)) +
  scale_fill_fermenter(
    type = "div",
    palette = "PuOr",
    n.breaks = 12,
    limits = c(-1, 1),
    name = "Effect Size",
    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
  ) +
  facet_wrap(~cluster, ncol = 1, strip.position = "right", scales = "free_y") +
  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
  xlab("377 pathways, ward.D2 clustered") +
  ylab("Clusters") +
  scale_x_discrete(position = "top") +
  guides(y.sec = guide_axis_label_trans(~ paste(.x))) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.ticks = element_blank(),
    strip.text.y = element_text(size = 10, hjust = .5),
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_blank(),
    panel.background = element_blank(),
    panel.spacing = unit(0, "lines"),
    legend.position = "bottom",
    legend.key.height = unit(0.3, "lines"),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 5),
    legend.margin=margin(0,0,0,0),
    legend.box.margin=margin(-10,-10,0,-10)
  )

load("../data/barplot_pathways_fig2b.rda")
fig2C <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
  geom_col() +
  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
  ylab("Proportion of mutated samples") +
  xlab("Significant pathways") +
  theme_classic() +
  scale_fill_manual(values = bp.cluster.colors) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    #strip.text.y = element_text(size = 10, hjust = .5),
    #plot.tag = element_text(size = 12, hjust = 0, vjust = 0),
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_blank(),
    legend.position = "none"
  ) #+ 
  ggtitle("Cluster-specific enrichment score")

layout <- '
AAABBBB
AAABBBB
AAABBBB
AAACCCC
AAACCCC
AAACCCC
AAACCCC
'

fig2A + fig2B + fig2C + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
```
